Tsbrowse: a genome browser for Ancestral Recombination Graphs

### 2025-03-07 manuscript/paper.tex

\section{Introduction} 
Ancestral recombination graphs (ARGs) describe how a set of sample
sequences relate to each other at each position along the genome in a recombining
species, and are currently the subject of intense research 
interest~\citep{brandt2024promise,lewanski2024era,nielsen2024inference,
wong2024general}. ARGs are a fundamental object in population genetics,
and although they have been of theoretical interest for
decades~\citep{Hudson1983,Griffiths1997} it is only with recent breakthroughs
in inference methods~\citep{rasmussen2014genome,
speidel2019method,kelleher2019inferring,wohns2022unified,zhang2023biobank,
gunnarsson2024scalable,deng2024robust} that practical application has 
become possible. Applications to varied questions such as the spatial 
location of genetic
ancestors~\citep{osmond2024estimating,deraje2024inferring,grundler2024geographic},

as well as the use of ARGs as training data for machine learning 
methods~\citep{whitehouse2024tree} depend critically on the \emph{accuracy}
of the inferred ARGs. Yet, despite this explosion of downstream methods using
inferred ARGs, there is a critical lack of practical tools to help
researchers make informed assesments about the quality of their inferred ARGs.

% Varied applications to
% simulation~\citep{kelleher2016efficient,baumdicker2021efficient,Anderson-Trocme2023}

Visualisation
is an essential component of data analysis pipelines. In the past few decades,
genome browsers have become synonymous with sequence visualisation, by
providing user-friendly, interactive interfaces to navigate genomic sequences
and annotations. From early standalone tools to manage \textit{C. elegans} data
\citep{Eeckman1995}, to modern web-apps for vertebrate genomes
\citep{Birney2004,Rangwala2024,Nassar2023}, browsers have continued to evolve
with the changing landscape of genomics. Modern genome browsers can load
datasets from a URL or local file system \citep{Lee2013,Robinson2023}.
The data security and scalability offered by this architecture ensures that
genome browsers continue to be relevant in the biobank era.

% Biobank-driven research has accelerated the development of several
% technologies, including the ability to infer Ancestral Recombination Graphs
% (ARGs) from large-scale genetic data
% % \citep{Kelleher2019,Speidel2019,Zhang2023,Gunnarsson2024.08.31.610248,Deng2024.03.16.585351}.
% ARGs encode the interwoven paths of genetic inheritance resulting from
% recombination, and can be represented as genetic genealogies along the sequence
% \citep{wong2024general,Hudson1983,Griffiths1997}. They capture rich detail about
% ancestral processes influencing the samples, and have garnered great interest
% in several applications in population and statistical genetics. These include
% the detection of ultra-rare genetic variation \citep{zhang2023biobank}, the mapping of
% quantitative trait loci \citep{Link2023}, and the sparse and efficient modeling
% of linkage disequilibrium \citep{Nowbandegani2023}. It is evident that ARG
% inference is increasingly becoming a part of genetic data analysis pipelines.
% Yet, ARG visualisation, which is crucial for downstream analyses remains
% challenging due to the complexity of inferences from empirical data. Existing
% visualisation tools (see \hyperref[sec:Discussion]{Discussion}) focus on the
% semantic representation of ARGs, but such representations have limited utility
% in informing downstream analyses.

To address this gap, we have developed \texttt{tsbrowse}, an open-source app
intended as a genome browser for ARGs. Its  focus is on visualising the
information structure in ARGs, thereby improving  inference and analysis at
biobank-scale. \texttt{Tsbrowse} provides interactive visual summaries of  the
fundamental properties of ARGs, i.e. the mutations, nodes, edges and trees they
encode. \texttt{Tsbrowse} operates on ARGs encoded as succinct tree sequences
\citep{wong2024general}. 
This the \texttt{tskit} library~\citep{ralph2020efficiently}
and supported by 
simulation~\citep{kelleher2016efficient,kelleher2018efficient,
haller2019tree,baumdicker2022efficient,
adrion2020community,lauterbur2023expanding,tsambos2023link,tagami2024tstrait}
inference~\citep{kelleher2019inferring,speidel2019method,wohns2022unified,
mahmoudi2022bayesian,zhan2023towards,zhang2023biobank,deng2025general}
and downstream processing methods~\citep{link2023tree,nowbandegani2023extremely}

@@

The interface is run on a web browser. It works on biobank-scale datasets
without requiring \textcolor{red}{raw data uploads to a server}. This is ideal
for performing quality control and exploratory analyses on ARGs inferred from
biobank datasets. \section{Results} \subsection{Architecture} \texttt{Tsbrowse}
is run in a step-wise approach (Figure \ref{fig:Figure_1}), which makes it
convenient to handle large datasets. The initial step (\texttt{tsbrowse
preprocess}) prepares the data for visualisation. It can be performed
separately on a large machine, and the output stored for later use. The second
step (\texttt{tsbrowse serve}), which can be run on a smaller machine, creates
and renders visualisations. This step follows a client-server architecture. On
the server side (the machine on which \texttt{tsbrowse serve} is run), the
Holoviz suite of libraries \citep{Holoviz} is leveraged to create
visualisations based on the pre-processed data. This choice of architecture is
prompted by the need to represent properties of large-scale ARGs in a
meaningful and performant way. For example, a visual representation of millions
of mutations in SARS-Cov-2 ARGs \citep{zhan2023towards} could quickly
lead to performance bottlenecks and overplotting. Datashader
\citep{Datashader}, a component Holoviz package, overcomes this challenge by
rasterising huge datasets (see \hyperref[subsec:User_Interface]{User Interface}
for details, and Supplementary Figure \ref{fig:Supplementary_Figure_4} for a
tsbrowse view of SARS-Cov-2 mutations). A web browser client connects to the
server to render and interact with these visualisations. Heavy data processing
is limited to the server side, keeping the client-side interface lightweight.
For large datasets with complex visualisations, this achieves a smoother user
experience.

>>>>

\section{Introduction} 
Ancestral recombination graphs (ARGs) describe how a set of sample
sequences relate to each other at each position along the genome in a recombining
species, and are currently the subject of intense research 
interest~\citep{brandt2024promise,lewanski2024era,nielsen2024inference,
wong2024general}. ARGs are a fundamental object in population genetics,
and although they have been of theoretical interest for
decades~\citep{Hudson1983,Griffiths1997} it is only with recent breakthroughs
in inference methods~\citep{rasmussen2014genome,
speidel2019method,kelleher2019inferring,wohns2022unified,zhang2023biobank,
gunnarsson2024scalable,deng2024robust} that practical application has 
become possible. Varied applications
such as inferring selection~\citep{stern2019approximate,hejase2022deep} and 
the spatial 
location of genetic
ancestors~\citep{osmond2024estimating,deraje2024inferring,grundler2024geographic},
more powerful approaches to quantifying genetic 
relatedness~\citep{fan2022genealogical,zhang2023biobank,gunnarsson2024scalable,lehmann2025on}
and other methodological improvements for genome wide association 
studies~\citep{nowbandegani2023extremely,link2023tree},
and the development of machine learning methods using inferred
ARGs as input~\citep{hejase2022deep,pearson2023local,korfmann2024simultaneous,
whitehouse2024tree} all depend critically on the \emph{accuracy}
of the inferred ARGs. 
Although studies benchmarking the various inference methods on 
simulated data have emerged 
\citep{brandt2022evaluation,deng2024robust,peng2024evaluating},
the practicalities of applying ARG inference to real data is 
understudied, and there is a critial lack of software infrastructure
to support evaluating the quality of estimated ARGs.

Visualisation is fundamentally important to data analysis.
Many specialised tools exists to aid the visual analysis and 
quality control
of assembly~\citep{wick2015bandage,challis2020blobtoolkit},
read mapping~\citep{robinson2011integrative},
and variant
calling~\citep{robinson2017variant,tollefson2019viva,konig2023divbrowse},
for example.
At every stage of a bioinformatics pipeline, it is important to visualise 
results to avoid artefacts and aid understanding of the data.
Genome browsers such as IGV~\citep{robinson2011integrative} and the 
UCSC Genome Browser~\citep{nassar2023ucsc} integrate many different
data modalities, and are vital infrastructure for the field.
% . In the past few decades,
% genome browsers have become synonymous with sequence visualisation, by
% providing user-friendly, interactive interfaces to navigate genomic sequences
% and annotations. From early standalone tools to manage \textit{C. elegans} data
% \citep{Eeckman1995}, to modern web-apps for vertebrate genomes
% \citep{Birney2004,Rangwala2024,Nassar2023}, browsers have continued to evolve
% with the changing landscape of genomics. Modern genome browsers can load
% datasets from a URL or local file system \citep{Lee2013,Robinson2023}.
% The data security and scalability offered by this architecture ensures that
% genome browsers continue to be relevant in the biobank era.

There is currently no straightforward means of visually summarising
ARGs, presenting a significant stumbling block for the nascent 
field of practical ARG inference. Inferred ARGs are essentially opaque,
with only the most basic numerical summaries (such as numbers 
of nodes, mutations etc) or high-level
statistics~\citep{ralph2020efficiently} available. While tools 
for visualising the local tree topologies exist, they are difficult
to interpret and do not scale well to large sample sizes. 
To address this gap, we present \texttt{tsbrowse}, a
% I worry that if we say "genome browser" we'll get stung for not having 
% deep integration with all sorts of stuff
client-server application providing genome browser-like 
functionality for ARGs. 
It provides interactive visualisations of the information structure 
of ARGs, smoothly scrolling from chromosome-level views down to 
individual nodes, edges and mutations. Supporting very large ARGs is a 
particular focus for \texttt{tsbrowse}, as millions of genome
sequences have been sampled for several
species~\citep{cesarani2022multibreed,stark2024call,hunt2024addressing}
and ARGs of this scale being actively
inferred~\citep{kelleher2019inferring,zhang2023biobank,
zhan2023towards,gunnarsson2024scalable}.

% Its focus is on interactively visualising the
% information structure in ARGs at all scales.

% \texttt{Tsbrowse} provides interactive visual summaries of  the
% fundamental properties of ARGs, i.e. the mutations, nodes, edges and trees they
% encode.

% \texttt{Tsbrowse} operates on ARGs encoded as succinct tree sequences
% \citep{wong2024general}. 
% This the \texttt{tskit} library~\citep{ralph2020efficiently}
% and supported by 
% simulation~\citep{kelleher2016efficient,kelleher2018efficient,
% haller2019tree,baumdicker2022efficient,
% adrion2020community,lauterbur2023expanding,korfmann2023weak,tsambos2023link,tagami2024tstrait}
% inference~\citep{kelleher2019inferring,speidel2019method,wohns2022unified,
% mahmoudi2022bayesian,zhan2023towards,zhang2023biobank,deng2025general}
% and downstream processing methods~\citep{fan2022genealogical,nowbandegani2023extremely}

@@

% The interface is run on a web browser. It works on biobank-scale datasets
% without requiring \textcolor{red}{raw data uploads to a server}. This is ideal
% for performing quality control and exploratory analyses on ARGs inferred from
% biobank datasets.

\section{Results} \subsection{Architecture} \texttt{Tsbrowse}
is run in a step-wise approach (Figure \ref{fig:Figure_1}), which makes it
convenient to handle large datasets. The initial step (\texttt{tsbrowse
preprocess}) prepares the data for visualisation. It can be performed
separately on a large machine, and the output stored for later use. The second
step (\texttt{tsbrowse serve}), which can be run on a smaller machine, creates
and renders visualisations. This step follows a client-server architecture. On
the server side (the machine on which \texttt{tsbrowse serve} is run), the
Holoviz suite of libraries \citep{Holoviz} is leveraged to create
visualisations based on the pre-processed data. This choice of architecture is
prompted by the need to represent properties of large-scale ARGs in a
meaningful and performant way. For example, a visual representation of millions
of mutations in SARS-Cov-2 ARGs \citep{zhan2023towards} could quickly
lead to performance bottlenecks and overplotting. Datashader
\citep{Datashader}, a component Holoviz package, overcomes this challenge by
rasterising huge datasets (see \hyperref[subsec:User_Interface]{User Interface}
for details, and Supplementary Figure \ref{fig:Supplementary_Figure_4} for a
tsbrowse view of SARS-Cov-2 mutations). A web browser client connects to the
server to render and interact with these visualisations. Heavy data processing
is limited to the server side, keeping the client-side interface lightweight.
For large datasets with complex visualisations, this achieves a smoother user
experience.
